They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-16 paper/paper.tex
tables 5 and 6, capital Ps

Chance (2020) re-examines the constant probability notion in Major
League Baseball's World Series (1923--2018), the National Basketball
Association's Finals (1951--2018), and the National Hockey League's
Stanley Cup (1939--2018). Chance finds strong evidence against the null
hypothesis of \(p = 0.5\) in the MLB and NHL championship series when
applying Mosteller's first and second methods.\footnote{Chance goes on
  to outline a conditional probability framework (likelihood of winning
  a game given the series state) which can exactly explain the
  distribution of the number of games played.} Chance's work is closely
related to mine, and, in fact, provides a guide for the first part of my
investigation.

@@

Studying both player streaks and team streaks in basketball, in both
observational and controlled settings. Gilovich et al. (1985) do not
find evidence for the hot hand phenomenon. Recently, Miller et al.
(2018) refuted the conclusions of Gilovich et al.~, finding mathematical
evidence that seems to support negative recency. Specifically, they find
that a ``bias exists in a common measure of the conditional dependence
of present outcomes on streaks of past outcomes in sequential data''
(streak selection bias) that imply that, under i.i.d. conditions, ``the
proportion of successes among the outcomes that immediately follow a
streak of consecutive successes is expected to be strictly less than the
underlying (conditional) probability of success''. I agree with Miller's
findings, accounting for the streak selection bias in my study of
momentum.

@@

\hypertarget{methodology-results-and-discussion}{%
\section{Methodology, results, and
discussion}\label{methodology-results-and-discussion}}

First, I investigate the constant probability assumption and the
distribution of rounds played in a series. Afterwards, I investigate
momentum, building on my learnings from the constant probability
assumption analysis.

@@

The general formula for the probability \(P_E(r)\) that a best-of-\(s\)
series lasts \(r\) rounds given constant probability \(p\) of one
team\footnote{If \(p > 0.5\), then I might say that this team is the
  better team (known omnisciently).} winning each round is

@@

\begin{longtable}{rrrr}
\caption{The probabilities that a best-of-11 series lasts $r$ rounds ($P_E(r)$, where $r \in [6, 7, 8, 9, 10, 11]$) under the assumption that each team has a 50\% probability ($p_0 = 0.5$) of winning each game. The observed frequencies observed are shown as a count $N_O(r)$ and as a proportion $P_O(r)$ of all series.}\label{tbl:cod-prob-series-lasting-r-rounds} \\
\toprule
&  & \multicolumn{2}{c}{Observed} \\ 
\cmidrule(lr){3-4}
Series lasts $r$ rounds & $P_E(r)$ & $P_O(r)$ & $N_O(r)$ \\ 
\midrule
6 & $3.1\%$ & $4.7\%$ & 40 \\ 
7 & $9.4\%$ & $11.9\%$ & 101 \\ 
8 & $16.4\%$ & $16.5\%$ & 141 \\ 
9 & $21.9\%$ & $21.7\%$ & 185 \\ 
10 & $24.6\%$ & $21.5\%$ & 183 \\ 
11 & $24.6\%$ & $23.7\%$ & 202 \\ 
\bottomrule
\end{longtable}

@@

Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new
\(P_E(r)\) when re-applying Equation \ref{eq:series-length} for the
maximum likelihood estimate \(p_2 = 0.575\).\footnote{The method of
  moments and minimum \(\chi^2\) estimates for \(p\) are omitted simply
  because the results would be nearly identical to those for the maximum
  likelihood estimate of \(p\) (since they are all \(\approx 0.575\).} I
observe that \(P_E(r)\) is notably larger for \(p_{2}\) when
\(r \in [6, 7]\), more closely matching \(P_O(r)\). \(p_{2}\) is also
closer to \(P_O(r)\) for \(r \in [9, 10]\), although not for
\(r \in [8, 11]\).

\begin{longtable}[]{@{}rrrr@{}}
\caption{The probabilities that a best-of-11 series lasts $r$ rounds under the naive assumption $p_0 = 0.5$ and the maximum likelihood estimation $p_2 = 0.575$.}\label{tbl:expected-series-lengths-alternative-ps} \\
\toprule()
Series lasts \(r\) rounds & \(p_0(r)\) = 0.5 & \(p_2(r)\) = 0.575 & \(P_O(r)\) \\
\midrule()
\endhead
6 & 3.1\% & 4.2\% & 4.7\% \\
7 & 9.4\% & 11.2\% & 11.9\% \\
8 & 16.4\% & 17.8\% & 16.5\% \\
9 & 21.9\% & 21.8\% & 21.7\% \\
10 & 24.6\% & 23.0\% & 21.5\% \\
11 & 24.6\% & 22.0\% & 23.7\% \\
\bottomrule()
\end{longtable}

@@

\begin{equation}\protect\hypertarget{eq:pwkr}{}{
  \hat{p}^{kr,\text{win}} = \hat{p}(\text{win round | k-round winning streak, series lasting r rounds})=\{
    \begin{array}{cl}
    \frac{6 - k}{i - k}, & \text{team wins series} \\
    \frac{11 - 6 - k}{i - k}, & \text{team loses series}
    \end{array}.
  }\label{eq:pwkr}
\end{equation}

\(\hat{p}^{kr,loss} = \hat{p}(\text{lose round | k-round losing streak, series lasting r rounds})\)
can be formulated symmetrically.

Miller et al. (2018) provide evidence for negative recency suggesting
that, given a deterministic i.i.d. Bernoulli probability (Equation
\ref{eq:pki},

\begin{equation}\protect\hypertarget{eq:pkrd}{}{
E[\hat{P}^{kr,\text{win}} - \hat{P}^{kr,\text{loss}}] = 0
}\label{eq:pkrd}\end{equation}

is not true due to streak-selection bias.\footnote{Note that these
  probabilities are separate from those in Section \ref{sec:analysis-1}.}
I don't attempt to measure this expectation like Miller et al. (2018),
who perform their analysis longitudinally per study participant;
nonetheless, I apply their formulation for applying the ``downward
bias'' on post-streak success to come up with estimates
\(\hat{P}_{kr,\text{win}}^{\text{adj.}}\), specifying \(r\) as the
number of trials.

Given that people typically perceive streaks as beginning after the
third success (or failure) at minimum (Carlson and Shu 2007), I focus on
round win percentages in rounds immediately following streaks of three
wins.\footnote{Three happens to also be a reasonable number for series
  that lasts at maximum 11 rounds.} Table
\ref{tbl:cod-round-win-prop-after-k-wins} shows
\(\hat{P}_{kr,\text{win}}\) for \(k = 3\) winning streaks, for both
teams that empirically won and lost the series. (Some combinations of
\(r\) and \(\text{team loses series}\) are irrelevant, such as the case
when \(r \in [7, 8, 9]\), so these combinations are excluded.)

\begin{longtable}{rrrrrrrr}
\toprule
&  &  & \multicolumn{3}{c}{$\hat{p}^{kr,\text{win}}$} & \multicolumn{2}{c}{$p$-value} \\
\cmidrule(lr){4-6} \cmidrule(lr){7-8}
Series lasts $r$ rounds & Win Series? & $N^{kr,\text{win}}_O$ & $O$ & $E$ & $\text{adj.}$ & $E$ & $\text{adj.}$ \\ 
\midrule
7 & win & 209 & $75\%$ & $75\%$ & $61\%$ & 0.58 & <=0.01 \\ 
8 & win & 209 & $62\%$ & $60\%$ & $45\%$ & 0.28 & <=0.01 \\ 
9 & win & 193 & $52\%$ & $50\%$ & $35\%$ & 0.33 & <=0.01 \\ 
10 & lose & 60 & $13\%$ & $14\%$ & $10\%$ & 0.64 & 0.26 \\ 
10 & win & 151 & $44\%$ & $43\%$ & $29\%$ & 0.45 & <=0.01 \\ 
11 & lose & 129 & $24\%$ & $25\%$ & $16\%$ & 0.63 & <=0.01 \\ 
11 & win & 150 & $40\%$ & $38\%$ & $25\%$ & 0.29 & <=0.01 \\ 
\bottomrule
\end{longtable}

@@

>>>>

Chance (2020) re-examines the constant probability notion in Major
League Baseball's World Series (1923--2018), the National Basketball
Association's Finals (1951--2018), and the National Hockey League's
Stanley Cup (1939--2018). Chance finds strong evidence against the null
hypothesis of \(p = 0.5\) in the MLB and NHL championship series when
applying Mosteller's first and second methods.{[}\^{}8{]} {[}\^{}8{]}:
Chance goes on to outline a conditional probability framework
(likelihood of winning a game given the series state) which can exactly
explain the distribution of the number of games played.

@@

Studying both player streaks and team streaks in basketball, in both
observational and controlled settings. Gilovich et al. (1985) do not
find evidence for the hot hand phenomenon. However, Miller and Sanjurjo
(2018) refuted the conclusions of Gilovich et al., proving a framework
for quantifying streak selection bias, which effectively works in the
manner posited by the gambler's fallacy\footnote{Streak selection bias,
  ``in conjunction with a quasi-Bayesian model of decision making under
  sample size neglect\ldots{} provides a novel structural candidate
  explanationfor the persistence of gambler's fallacy beliefs'', per
  Miller and Sanjurjo (2018).}. Specifically, Miller and Sanjurjo say
that a ``bias exists in a common measure of the conditional dependence
of present outcomes on streaks of past outcomes in sequential data''
implying that, under i.i.d. conditions, ``the proportion of successes
among the outcomes that immediately follow a streak of consecutive
successes is expected to be strictly less than the underlying
(conditional) probability of success''. When applying streak selection
bias to Giloveih et al.'s data, Miller and Sanjurjo come to the opposite
conclusions as Giloveih et al.

@@

\hypertarget{methodology-and-results}{%
\section{Methodology and results}\label{methodology-and-results}}

First, I investigate the constant probability assumption and the
distribution of rounds played in a series. Chance's (2020) work is
closely related to mine, and, in fact, provides a guide for this
investigation.

Afterwards, I investigate post-streak win rates, building on my
learnings from the constant probability assumption analysis. I agree
with Miller and Sanjurjo's (2018) framework for streak selection bias,
accounting for it in when analyzing momentum.

@@

The general formula for the probability \(P_E(r)\) that a best-of-\(s\)
series lasts \(r\) rounds given constant probability \(p\) of one
team\footnote{If \(p > 0.5\), then one might say that this team is the
  better team (known in hindsight).} winning each round is

@@

\begin{longtable}{rrrr}
\caption{The probabilities that a best-of-11 series lasts $r$ rounds ($P_E(r)$, where $r \in [6, 7, 8, 9, 10, 11]$) under the assumption that each team has a 50\% probability ($p_0 = 0.5$) of winning each game. The observed frequencies observed are shown as a count $N_O(r)$ and as a proportion $P_O(r)$ of all series.}\label{tbl:cod-prob-series-lasting-r-rounds} \\
\toprule
&  & \multicolumn{2}{c}{Observed} \\ 
\cmidrule(lr){3-4}
$r$ & $P_E(r)$ & $P_O(r)$ & $N_O(r)$ \\ 
\midrule
6 & $3.1\%$ & $4.7\%$ & 40 \\ 
7 & $9.4\%$ & $11.9\%$ & 101 \\ 
8 & $16.4\%$ & $16.5\%$ & 141 \\ 
9 & $21.9\%$ & $21.7\%$ & 185 \\ 
10 & $24.6\%$ & $21.5\%$ & 183 \\ 
11 & $24.6\%$ & $23.7\%$ & 202 \\ 
\bottomrule
\end{longtable}

@@

Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new
\(P_E(r)\) when re-applying Equation \ref{eq:series-length} for the
maximum likelihood estimate \(p_2 = 0.575\).\footnote{The method of
  moments and minimum \(\chi^2\) estimates for \(p\) are omitted simply
  because the results would be nearly identical to those for the maximum
  likelihood estimate of \(p\) (since they are all \(\approx 0.575\)).}
I observe that \(P_E(r)\) is notably larger for \(p_{2}\) when
\(r \in [6, 7]\), more closely matching \(P_O(r)\). \(p_{2}\) is also
closer to \(P_O(r)\) for \(r \in [9, 10]\), although not for
\(r \in [8, 11]\).

\begin{longtable}[]{@{}rrrr@{}}
\caption{The probabilities that a best-of-11 series lasts $r$ rounds under the naive assumption $p_0 = 0.5$ and the maximum likelihood estimation $p_2 = 0.575$.}\label{tbl:expected-series-lengths-alternative-ps} \\
\toprule()
$r$ & $p_0(r)$ = 0.5 & $p_2(r)$ = 0.575 & $P_O(r)$ \\
\midrule()
\endhead
6 & 3.1\% & 4.2\% & 4.7\% \\
7 & 9.4\% & 11.2\% & 11.9\% \\
8 & 16.4\% & 17.8\% & 16.5\% \\
9 & 21.9\% & 21.8\% & 21.7\% \\
10 & 24.6\% & 23.0\% & 21.5\% \\
11 & 24.6\% & 22.0\% & 23.7\% \\
\bottomrule()
\end{longtable}

@@

\begin{equation}\protect\hypertarget{eq:pwkr}{}{
  \hat{P}^{kr,\text{win}}_{E,\text{unadj.}} = \hat{P}(\text{win round | k-round winning streak, series lasting r rounds})=\{
    \begin{array}{cl}
    \frac{6 - k}{i - k}, & \text{team wins series} \\
    \frac{11 - 6 - k}{i - k}, & \text{team loses series}
    \end{array}.
  }\label{eq:pwkr}
\end{equation}

\(\hat{P}^{kr,loss}_{E,\text{unadj.}} = \hat{P}(\text{lose round | k-round losing streak, series lasting r rounds})\)
can be formulated symmetrically.

While one might think that
\(E[\hat{P}^{kr,\text{win}} - \hat{P}^{kr,\text{loss}}] = 0\), Miller
and Sanjurjo (2018) suggest that, given a deterministic i.i.d. Bernoulli
probability (Equation \ref{eq:pwkr}, this is not true due to
streak-selection bias.\footnote{Note that these probabilities are
  separate from those in Section \ref{sec:analysis-1}.} As proved by
Miller and Sanjurjo, TODO: Combine these two sentences

\begin{quote}
Theorem 1: \emph{Let}
\(\mathbf{X} = \{X_i\}^{n}_{i=1}, n \geq 3\)\emph{, be a sequence of
independent Bernoulli trials, each with probability of success}
\(0 < p < 1\)\emph{. Let} \(\hat{P}_k(\mathbf{X})\) \emph{be the
proportion of successes on the subset of trials} \(I_k(\mathbf{X})\)
\emph{that immediately follow} \(k\) consecutive successes, that is,
\(\hat{P}_k(\mathbf{X}) := \sum_{i \in I_k(\mathbf{X})} X_i | I_k(\mathbf{X}) |. \hat{P}_k\)
\emph{is a biased estimate of}
\(\mathbb{P}(X_t = 1 | \prod_{j=t-k}^{t-1} X_j = 1) \equiv p\) \emph{for
all} \(k\) \emph{such that} \(1 \leq k \leq n - 2\)\emph{. In
particular,}
\(E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset] < p.\)
\end{quote}

TODO: Descripe the Appendix E.1. I apply their formulation for
``downward bias'' to post-streak success to come up with estimates
\(\hat{P}^{kr,\text{win}}_{E,\text{adj.}}\), specifying \(r\) as the
number of trials.

Given that people typically perceive streaks as beginning after the
third success (or failure) at minimum (Carlson and Shu 2007), I focus on
round win percentages in rounds immediately following streaks of three
wins.\footnote{Three happens to also be a reasonable number for series
  that lasts at maximum 11 rounds.} Table
\ref{tbl:cod-round-win-prop-after-k-wins} and table
\ref{tbl:cod-round-loss-prop-after-k-losses} shows
\(\hat{P}^{kr,\text{win}}_O\) and \(\hat{P}^{kr,\text{loss}}_O\) for
\(k = 3\) streaks respectively.

TODO: Put p-value in parentheses

\begin{longtable}{rrrrrrr}
\caption{Observed proportion of wins ($\hat{P}^{kr,\text{win}}_O$) and expected proportion of wins, both un-adjusted ($\hat{P}^{kr,\text{win}}_{E,\text{unadj.}}$) and adjusted ($\hat{P}^{kr,\text{win}}_{E,\text{adj.}}$), after 3-round-win streak, for a series lasting $r$ rounds.}\label{tbl:cod-round-win-prop-after-k-wins} \\
\toprule
 &  &  & \multicolumn{2}{c}{\text{Unadjusted}} & \multicolumn{2}{c}{\text{Adjusted}} \\ 
\cmidrule(lr){4-5} \cmidrule(lr){6-7}
$r$ & $N^{kr,\text{win}}_O$ & $\hat{P}^{kr,\text{win}}_O$ & $\hat{P}^{kr,\text{win}}_{E,\text{unadj.}}$ &  $p$-value & $\hat{P}^{kr,\text{win}}_{E,\text{adj.}}$ & $p$-value \\ 
\midrule
7 & 209 & $75\%$ & $75\%$ & 0.58 & $61\%$ & <=0.01 \\ 
8 & 209 & $62\%$ & $60\%$ & 0.28 & $45\%$ & <=0.01 \\ 
9 & 193 & $52\%$ & $50\%$ & 0.33 & $35\%$ & <=0.01 \\ 
10 & 151 & $44\%$ & $43\%$ & 0.45 & $29\%$ & <=0.01 \\ 
11 & 150 & $40\%$ & $38\%$ & 0.29 & $25\%$ & <=0.01 \\ 
\bottomrule
\end{longtable}

\begin{longtable}{rrrrrrr}
\caption{Observed proportion of losses ($\hat{P}^{kr,\text{loss}}_O$) and expected proportion of losses, both un-adjusted ($\hat{P}^{kr,\text{loss}}_{E,\text{unadj.}}$) and adjusted ($\hat{P}^{kr,\text{loss}}_{E,\text{adj.}}$), after 3-round-loss streak, for a series lasting $r$ rounds. $r \in [7, 8, 9]$ are invalid.}\label{tbl:cod-round-loss-prop-after-k-losses} \\
\toprule
 &  &  & \multicolumn{2}{c}{\text{Unadjusted}} & \multicolumn{2}{c}{\text{Adjusted}} \\ 
\cmidrule(lr){4-5} \cmidrule(lr){6-7}
$r$ & $N^{kr,\text{loss}}_O$ & $\hat{P}^{kr,\text{loss}}_O$ & $\hat{P}^{kr,\text{loss}}_{E,\text{unadj.}}$ &  $p$-value & $\hat{P}^{kr,\text{loss}}_{E,\text{adj.}}$ & $p$-value \\ 
\midrule
10 & 60 & $13\%$ & $14\%$ & 0.64 & $10\%$ & 0.26 \\ 
11 & 129 & $24\%$ & $25\%$ & 0.63 & $16\%$ & <=0.01 \\ 
\bottomrule
\end{longtable}

@@

\hypertarget{appendix}{%
\section*{Appendix}\label{appendix}}
\addcontentsline{toc}{section}{Appendix}
